\(Q\)-Bosons and their relations with integrable hierarchies

### 2024-04-16 draft.tex
finish phase model

@@

\subsubsection{Schur functions expansion for other correlation functions}
More generally, let us consider the correlation functions 
\begin{equation}
\begin{split}
  P_{\lambda_1 \lambda_2}(N', N, M|\bm{x}, \bm{y}) & =
  \bra{\lambda_1} \prod_{j=1}^N \mathbb{C}(x_j)
  \prod_{k=1}^{N'} \mathbb{B}(y_k)\ket{\lambda_2} \\
  & = \sum_{\mu} s_{\mu/\lambda_1}(\bm{x}) s_{\mu/\lambda_2}(\bm{y}) \; .
\end{split}
\end{equation}
where \((N' + \ell(\lambda_2) + n_0 = N' + \ell(\lambda_2) + n'_0)\) and 
we have also used that the expression for any Young diagram
\(\mu\) that does not contain \(\lambda_1\) and \(\lambda_2\)
vanishes. Using the elementary property of skew Schur
polynomials~\cite{Macdonald:1998}
\begin{equation}
  \sum_{\mu} s_{\mu/\lambda_1}(\bm{x}) s_{\mu/\lambda_2}(\bm{y}) = \prod_{i,j}\frac{1}{1 - x_i y_j}
 \sum_\nu s_{\lambda_1/\nu}(\bm{x}) s_{\lambda_2/\nu}(\bm{y})\; ,
\end{equation}
and the Cauchy's identity, we have 
\begin{equation}
\begin{split}
  P_{\lambda_1 \lambda_2}(N', N, M|\bm{x}, \bm{y}) & =
 \sum_\mu s_{\mu}(\bm{x}) s_{\mu}(\bm{y})
 \sum_\nu s_{\lambda_1/\nu}(\bm{x}) s_{\lambda_2/\nu}(\bm{y})\; .
\end{split}
\end{equation}
Therefore, this correlation function is nothing but the off-shell
norm~(\ref{eq:scalar}) times a finite sum over skew Schur functions.
This expression shows that the existence of these tau functions is not
a trivial fact of this model.

In fact, it is worth considering other quantities that are not
necessarily tau functions. Consider the state
\begin{equation}
  \ket{\mathcal{Y}} =
 \sum_{\vec{n}} \prod_{j=0}^M \ket{n_j}\quad \sum_j n_j = N\; .
\end{equation}
We have~\cite{Bogoliubov2005} the correlation function 
\begin{equation}
  \prod_{j=1}^N x^{M/2} \bra{\mathcal{Y}} \prod_{j=1}^N B(\bm{x}) \ket{\bm{0}} = 
  \sum_{\lambda\subseteq (N,N,M)} s_\lambda(\bm{x})\; . 
\end{equation}
This sum of Schur polynomials are not tau functions of the KP hierarchy.

We can easily see this fact, if we consider the the Fermionic
construction. Since
\begin{equation}
\tau(\bm{t}) = \bra{\bm{0}} e^{\bm{J}_+(\bm{t})} G \ket{\bm{0}}
\end{equation}
where \(G\in GL(\infty)\), and 
\begin{equation}
  \bra{\bm{0}} e^{\bm{J}_+(\bm{t})} = \sum_\lambda \bra{\mu} s_\lambda(\bm{t})\; , 
\end{equation}
therefore
\begin{equation}
\tau(\bm{t}) = \sum_{\lambda} \bra{\lambda} G \ket{\bm{0}} s_\lambda(\bm{t})
\end{equation}
and we need \(1 = \bra{\lambda} G \ket{\bm{0}} \). Then, 
\begin{equation}
  \ket{\Psi_0} = G\ket{\bm{0}} = \sum_\lambda \ket{\lambda}\; ,
\end{equation}
is a coherent state, and we conclude that
\begin{equation}
G = 1 +  \sum_{d\geq 1} \prod_{j=1}^d \psi^\star_{a_j} \psi_{-b_j}\ \notin \ GL(\infty) \;. 
\end{equation}
It is interesting to notice that this state appears as the ground
state of the crystal melting problem in~\cite{Dijkgraaf:2008ua}.

Therefore, the restriction for diagrams with \(\ell(\lambda_1) \leq N\) is
not a tau function either.

>>>>

\subsubsection{Skew Schur polynomials expansion}
More generally, let us consider the correlation functions 
\begin{equation}
\begin{split}
  P_{\lambda_1 \lambda_2}(N', N, M|\bm{x}, \bm{y}) & =
  \bra{\lambda_1} \prod_{j=1}^N \mathbb{C}(x_j)
  \prod_{k=1}^{N'} \mathbb{B}(y_k)\ket{\lambda_2} \\
  & = \sum_{\mu} s_{\mu/\lambda_1}(\bm{x}) s_{\mu/\lambda_2}(\bm{y}) \; .
\end{split}
\end{equation}
where \(N' + \ell(\lambda_2) + n_0 = N' + \ell(\lambda_2) + n'_0\) and 
we have also used that the expression for any Young diagram
\(\mu\) that does not contain \(\lambda_1\) and \(\lambda_2\)
vanishes. Using the elementary property of skew Schur
polynomials~\cite{Macdonald:1998}
\begin{equation}
  \sum_{\mu} s_{\mu/\lambda_1}(\bm{x}) s_{\mu/\lambda_2}(\bm{y}) = \prod_{i,j}\frac{1}{1 - x_i y_j}
 \sum_\nu s_{\lambda_1/\nu}(\bm{x}) s_{\lambda_2/\nu}(\bm{y})\; ,
\end{equation}
and the Cauchy's identity, we have 
\begin{equation}
\begin{split}
  P_{\lambda_1 \lambda_2}(N', N, M|\bm{x}, \bm{y}) & =
 \sum_\mu s_{\mu}(\bm{x}) s_{\mu}(\bm{y})
 \sum_\nu s_{\lambda_1/\nu}(\bm{x}) s_{\lambda_2/\nu}(\bm{y})\; .
\end{split}
\end{equation}
Therefore, this correlation function is nothing but the off-shell
norm~(\ref{eq:scalar}) times a finite sum over skew Schur functions.
This expression shows that the existence of these tau functions is not
a trivial fact of this model.

@@

\subsection{More tau functions}
Based on the discussion we had so far, and on the general map between 
the Phase model and the free fermions, one can understand the general form 
of tau functions in this phase model context.

Let us consider the vertex operator construction~\cite{Okounkov2001},
see also~\cite{Alexandrov:2012tr, Wheeler:2010vmq}. Consider the
vacuum \(\ket{\bm{0}}\), a ``Dirac sea'', defined by the conditions \(\psi_m\ket{\bm{0}}
= \psi_n^\star\ket{\bm{0}} =0 \) for \(m<0\) and \(n \geq 0\), where 
\(\psi_n\) are components of a holomorphic free fermionic field. 
In this formalism, the partition states are given by 
\begin{equation}
  \ket{\mu} = \textrm{sign}(\sigma) \prod_{j=1}^d \psi_{a_j} \psi^\star_{-b_j}\ket{\bm{0}}\; ,
\end{equation}
where \(\textrm{sign}(\sigma) = \pm 1\) are defined in a such a way
that the Shur coefficients of the vertex operators, defined below,
have positive coefficients.

The pairs \((a_j|b_j)_{j=1}^d\) define the Frobenius notation of the
partition \(\mu = (\mu_1, \mu_2, \dots, \mu_\ell)\), that is given by
\(a_j = \mu_j - j\) and \(b_j = \mu^T_j - j\), where \(d\) is the
number of boxes in the diagonal of the Young diagram, and \(\mu^T\) is
its conjugate, or transpose.  From these definitions, we have the
equivalence
\begin{equation}
  \prod_{k\geq 1} (\phi_k^\dagger)^{n_k} \mapsto \textrm{sign}(\sigma) \prod_{j=1}^d
  \psi_{a_j} \psi^\star_{-b_j} \qquad |\mu| = \sum_k k n_k = \sum_j(a_j + b_j) + d\; .
\end{equation}

Finally, the \(\mathfrak{gl}(\infty)\) has generators given by the
bilinears \(X = \sum_{j, j \in \mathbb{Z}} x_{ij} \bm{\colon} \psi_i \psi_j^\star\bm{\colon}
+ c\), where \(c\in \mathbb{C}\), \(x_{ij} =
0\) for large \(|j -i|\), say \(\geq M\), and the colons denote the normal ordering
\begin{equation}
  \bm{\colon} \psi_i \psi_j^\ast \bm{\colon} =  \psi_i \psi_j^\ast
  - \bra{0} \psi_i \psi_j^\ast \ket{0} \; .
\end{equation}
The group \(GL(\infty)\) is defined through the \(\exp:
\mathfrak{gl}(\infty) \to GL(\infty)\) as usual.

Observe that these elements have finitely many non zero elements, and
that these include diagonal terms \(\bm{\colon} \psi_i
\psi_i^\star\bm{\colon}\), upper and lower triangular terms. These are
equivalent to the Number operators \(\mathcal{N}\), annihilation
\(\phi\) and creation \(\phi^\dagger\) , respectivelly.  The central
charge is equivalent to the vaccum projection \(\pi =
\ket{\bm{0}}\bra{\bm{0}}\). Therefore, the elements
\begin{equation}
  G \in GL(\infty) \mapsto
  \mathcal{G} = \exp \left(\sum_{i=1}^M \sum_{a=1}^3  c_{i, a} T_i^{a}  + c \pi_i\right)
\end{equation}
\(T^{(a)}_i \in \{ \mathcal{N}_i, \phi_i, \phi_i^\dagger \}_{i=1}^M\).

The operators \(\mathbb{B}(x)\) and \(\mathbb{C}(x)\), 
for a large enough chain \(M\to \infty\),
are related to the vertex operators
\(\Gamma_-(x)\) and \(\Gamma_+(x)\), respectivelly, as
\begin{equation}
  \begin{split}
    \mathbb{B}(x) \mapsto \Gamma_-(x) & = \exp \left( \sum_{n\geq 1} \frac{1}{n}x^n J_{-n}\right) \\
    \mathbb{C}(x) \mapsto \Gamma_+(x) & = \exp \left( \sum_{n\geq 1} \frac{1}{n}x^n J_{n}\right) \; ,
  \end{split}
\end{equation}
where \(J_n \) is written, in terms of free fermions, as \(J_n =
\sum_{j\in \mathbb{Z}} \bm{\colon} \psi_j \psi_{j+n}^\ast
\bm{\colon}\),
These set of operators generate a Heisenberg subalgebra
\(\widehat{\mathfrak{gl}}(1) \subset \mathfrak{gl}(\infty)\) given by
\begin{equation}
  [J_m, J_n] = m \delta_{n+m,0}\; .
\end{equation}

Putting all these facts together, we have that the tau functions of the Toda 
hierarchies, given by
\begin{equation}
  \tau_s(\bm{x}, \bm{y}) = \bra{s} \prod_{i} \Gamma_+(x_i) G \prod_j \Gamma_+(x_j) \ket{s}\; ,
\end{equation}
are maped into objects of the form 
\begin{equation}
\begin{split}
  \tau_0(\bm{x}, \bm{y}) & = \bra{\Psi(\bm{x})} \mathcal{G} \ket{\Psi(\bm{y})} \\
  \\ & = \bra{\bm{0}} \prod_{i} \mathbb{C}(x_i)
  \mathcal{G} \prod_j \mathbb{B}(y_j) \ket{\bm{0}}\; ,
\end{split}
\end{equation}
where we necessarily have \(s=0\) in the phase model

\subsubsection{Schur polynomials expansion of some correlation functions}
In fact, it is worth noticing that other quantities, that are not tau
functions, might have interesting Schur polynomials expansion.
Consider the state calculated in~\cite{Bogoliubov2005}
\begin{equation}
  \ket{\mathcal{Y}} =
 \sum_{\vec{n}} \prod_{j=0}^M \ket{n_j} = \sum_\mu \ket{\mu} \qquad \sum_j n_j = N\; .
\end{equation}
Then
\begin{equation}
 \bra{\mathcal{Y}} \prod_{j=1}^N B(\bm{x}) \ket{\nu} = 
  \sum_{\lambda\subseteq (N,N,M)} s_{\lambda/\nu}(\bm{x})\; . 
\end{equation}
And it is just a sum of skew Schur polynomials.
